---
title: 'Simulation Stimulation Part 2: AFL Season 2019'
author: Joshua White
date: '2019-03-30'
slug: simulation-stimulation-part-2-afl-season-2019
categories:
  - R
tags:
  - AFL
  - Simulations
  - Sport
image:
  caption: ''
  focal_point: ''
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("blogfiles/AFL-Sim/AFLsim2019.R")
```

In part one of this series, I applied three different models to the 2018 AFL season. In this part, I apply these models to the 2019 AFL fixture to attempt to make some predictions about the standings at the end of the year. 

Now Ideally, I would have applied these models to the last 10 (or more) years of AFL data to see which one better predicts results for the upcoming year --- but who has time for that --- So we are just going to go ahead. 

## The Models: A Reminder

###Poisson Distributed Team Scores
Here, each team's score in every game is modeled as a random drawing from a Poisson distribution --- i.e. $P(k) = e^{-\lambda} \frac{\lambda^k}{k!}$, where k is the number of points scored, and $\lambda$ is calculated as 

$\lambda_{home team} =$ League Average points scored at home $\times$ home team attack strength at home $\times$ awayteam attack strength away

$\lambda_{away team} =$ League Average points scored away $\times$ away team attack strength away $\times$ hometeam defense strength at home


###Normally Distributed team scores
This model assumes that team scores are normally distributed around $\lambda$ as defined above, with a standard deviation calculated as the pooled standard deviation of that teams 'for' scores standard deviation, and the opponents 'against' scores standard deviation:
$$Points_{teamA} \sim \mathcal{N}(\mu=\lambda,\,\sigma = \sigma_p)\,$$ 
where: <br>
$$\sigma_{p}=\frac{\sigma_{teamAfor} + \sigma_{teamBagainst}}{2}$$

###Rankings model: tanh
This model simply relies on the ordinal data of last years final Ladder: Carlton, the wooden spoon winners (i.e. the last placeholders, out of 18 teams) get a ranking of 1, and Richmond, who finished on top of the table get a ranking of 18. Then, with this information, the probability of team A winning is modelled as follows:

$$P(A\>wins)=\frac{tanh(\frac{a-b}{w} + d)+1}{2}$$
where,<br>
A = home team <br>
a = home team ranking <br> 
b = away team ranking <br>
w = weighting parameter (the higher the number the less weight placed on the rankings) <br>
d = draw paramater, to allow for possibility of draws, calculated as so that $tanh(d) = -draw\>rate$

## 2019 Results

###Poisson model
```{r poisresults}
Finalsimtablepois.2019
pois2019graph
```

###Normal model
```{r normresults}
Finalsimtablenorm.2019
norm2019graph
```

###Tanh model

```{r tanhresults}
Finalsimlist.2019[[20]] %>% tibble()
tanh2019graph
```


##Comparing the three models.

```{r comparing the models, eval=FALSE}

#create table with different orders. 
ordercomparison.2018 <- data.frame(Team = factor(finalTable2018$Team),
                              Order2018 = 1:18,
                              OrderNorm.2018 = match(finalTable2018$Team, Finalsimtablenorm.2018$Team),
                              OrderPois.2018 = match(finalTable2018$Team, Finalsimtablepois.2018$Team),
                              OrderTanh.2018 = match(finalTable2018$Team, Finalsimlist.2018[[20]]$Team)) 

ordercomparison.2018 %>% tibble()

#Let's now get the spearman's correlation coefficients and show that in a correlatin matrix
cormatrix.2018 <- cor(ordercomparison.2018[,2:5], use = "all.obs", method = "spearman")
ggcorrplot(cormatrix.2018, lab = TRUE)

## create comparison graphs ##

#combine two graphs##
Finalsimtablenorm.2019$Model <- "norm"
Finalsimtablepois.2019$Model <- "pois"
Finalsimlist.2019[[20]]$Model <- "tanh"

comparisonSimTable.2019 <- rbind(Finalsimtablenorm.2019, Finalsimtablepois.2019, Finalsimlist.2019[[20]])
comparisonSimTable.2019$Model <- factor(comparisonSimTable.2019$Model)

# create tidy data to compare a lot in one go
comparisonSimTable.2019.tidy <- gather(comparisonSimTable.2019, "Type", "Percent", 6:10)
comparisonSimTable.2019.tidy 

#create ordered factor to order graph
comparisonSimTable.2019.tidy$Team <- factor(comparisonSimTable.2019.tidy$Team, levels = Finalsimtablepois.2019$Team)

#create ordered factor of 'Type' to order facet
comparisonSimTable.2019.tidy$Type <- factor(comparisonSimTable.2019.tidy$Type, 
                                       levels = c("percentWon", "percentTop4", "percentTop8",
                                                  "percentBottom4", "percentLast"))

#draw graphs
ggplot(comparisonSimTable.2019, aes(x = Team, y = percentWon, fill = Model)) + 
  geom_bar(stat = "identity", position = 'dodge')
ggplot(comparisonSimTable.2019, aes(x = Team, y = percentTop8, fill = Model)) +
  geom_bar(stat = "identity", position = 'dodge')

ggplot(comparisonSimTable.2019.tidy, aes(x = Team, y = Percent, fill = Model)) + 
  geom_bar(stat = "identity", position = 'dodge') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  facet_wrap(~Type, nrow = 1)

```
